\pkg{depmixS4} : An \proglang{R}-package for hidden Markov models

### 2010-02-24 papers/jss/dpx4Rev.Rnw
Added notes on balance and speed data sets.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@367 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

%The maximising values of $a_{jk}$ are \cite<e.g.,>{Rabiner89}
%\begin{equation}
%\label{eq:MLA}
%\hat{a}_{jk} = \frac{1}{N(T-1)} \sum_{i=1}^N \sum_{t=2}^T  \frac{\xi^i_t(j,k)}{\gamma_{t-1}^i(j)}.
%\end{equation}
%Fixing certain elements of $\mat{A}$ to 0, as in the DT version, does not affect the estimation of the other elements. When elements $a_{jk}$ are assumed identical, we simply extend the summation in (\ref{eq:MLA}) to include all those elements, changing the denominator $N(T-1)$ accordingly.

%To estimate $\greekv{\lambda}$, we note that the term containing this parameter has the form of a weighted likelihood, where $\gamma_t^i(j)$ can be interpreted as the number of replications of $r^i_t$. Hence, we can rely on standard ML estimates of the logistic regression coefficients $\lambda_i$, using the values $\gamma_t^i(j)$ as ``case weights'' \cite<e.g.,>{Agresti02,McCullagh83}\footnote{To be more specific, we replicate each observation $r^i_t$ a total of 6 times, once for each of the states besides the random state (which offers no information regarding $\lambda_i$). For the $j$-th replication (corresponding to the $j$-th state), we used $v_j(\vec{x}_t)$ as a predictor variable and $\gamma^i_t(j)$ as a case weight. All these replications were used to obtain the maximum likelihood estimate of $\lambda$ from a single GLM, using the ``glm.fit'' function in R \cite{R}.}. The resulting parameter estimates are then used in a new Expectation step, followed by another Maximisation step, until convergence. Under mild regularity conditions, it can be shown that the EM algorithm converges to a (local) maximum of the likelihood \cite{EM}.

@@

Two steps are involved in using \pkg{depmixS4} which are illustrated
below with examples:
\begin{enumerate}
	\item model specification with function \code{depmix} (or with \code{mix}
	for latent class and finite mixture models, see example below on adding
	covariates to prior probabilities)

@@

Throughout this article a data set called \code{speed} is used.  As
already indicated in the Introduction, it consists of three time
series with three variables: response time, accuracy, and a covariate
Pacc which defines the relative pay-off for speeded versus accurate
responding.  The participant in this experiment switches between fast
responding at chance level and relatively slower responding at a high
level of accuracy.  Interesting hypotheses to test are: is the
switching regime symmetric?  Is there evidence for two states or does
one state suffice?  Is the guessing state actually a guessing state,
i.e., is the probability of a correct response at chance level (0.5)?

@@

A dependent mixture model is defined by the number of states and the
initial state, state transition, and response distribution functions.
A dependent mixture model can be created with the
\code{depmix}-function as follows:

<<>>=
library(depmixS4)
data(speed)
set.seed(1)
mod <- depmix(rt~1, data=speed, nstates=2, trstart=runif(4))
@

The \code{depmix} function returns an object of class \code{depmix}
which contains the model specification (and not a fitted model!).
Note also that start values for the transition parameters are provided
in this call using the \code{trstart} argument.  At this time, the
package does not provide automatic starting values.

@@

The \code{summary} method of \code{fit}ted models provides the
parameter estimates, first for the prior probabilities model, second
for the transition model, and third for the response models.
<<>>=
summary(fm)
@

Note that, since no further arguments were specified, the initial
state, state transition and response distributions were set to their
defaults (multinomial distributions for the first two, and Gaussian
distributions for the response distributions).

\subsection{Covariates on transition parameters}

@@

\subsection{Adding covariates on the prior probabilities}

To illustrate the use of covariates on the prior probabilities we have
included another data set with \pkg{depmixS4}.  The \code{balance}
data consists of 4 binary items (correct-incorrect) on a balance scale
task \citet{Siegler1981}.  The data form a subset of the data
published in \citet{Jansen2002}.

>>>>

%The maximising values of $a_{jk}$ are \cite<e.g.,>{Rabiner89}
%\begin{equation}
%\label{eq:MLA}
%\hat{a}_{jk} = \frac{1}{N(T-1)} \sum_{i=1}^N \sum_{t=2}^T  \frac{\xi^i_t(j,k)}{\gamma_{t-1}^i(j)}.
%\end{equation}

%Fixing certain elements of $\mat{A}$ to 0, as in the DT version, does
%not affect the estimation of the other elements.  When elements
%$a_{jk}$ are assumed identical, we simply extend the summation in
%(\ref{eq:MLA}) to include all those elements, changing the
%denominator $N(T-1)$ accordingly.

%To estimate $\greekv{\lambda}$, we note that the term containing this
%parameter has the form of a weighted likelihood, where
%$\gamma_t^i(j)$ can be interpreted as the number of replications of
%$r^i_t$.  Hence, we can rely on standard ML estimates of the logistic
%regression coefficients $\lambda_i$, using the values $\gamma_t^i(j)$
%as ``case weights'' \cite<e.g.,>{Agresti02,McCullagh83}\footnote{To
%be more specific, we replicate each observation $r^i_t$ a total of 6
%times, once for each of the states besides the random state (which
%offers no information regarding $\lambda_i$).  For the $j$-th
%replication (corresponding to the $j$-th state), we used
%$v_j(\vec{x}_t)$ as a predictor variable and $\gamma^i_t(j)$ as a
%case weight.  All these replications were used to obtain the maximum
%likelihood estimate of $\lambda$ from a single GLM, using the
%``glm.fit'' function in R \cite{R}.}.  The resulting parameter
%estimates are then used in a new Expectation step, followed by
%another Maximisation step, until convergence.  Under mild regularity
%conditions, it can be shown that the EM algorithm converges to a
%(local) maximum of the likelihood \cite{EM}.

@@

Two steps are involved in using \pkg{depmixS4} which are illustrated
below with examples:
\begin{enumerate}
	\item model specification with function \code{depmix} (or with \code{mix}
	for latent class and finite mixture models, see example below on adding
	covariates to prior probabilities in section~\ref{sec:prior})

@@

Throughout this article a data set called \code{speed} is used.  As
already indicated in the Introduction, it consists of three time
series with three variables: response time, accuracy, and a covariate,
Pacc, which defines the relative pay-off for speeded versus accurate
responding.  Before describing some of the models that are fitted
to these data, we provide a brief sketch of the reasons for gathering
these data in the first place.

Response times are a very common dependent variable in psychological
experiments and hence form the basis for infernce about many
psychological processes.  A potential threat to such inference based
on response times is formed by the speed-accuracy trade-off: different
participants in an experiment may respond differently to typical
instructions to `respond as fast and accurate as possible'.  A popular
model which takes the speed-accuracy trade-off into account is the
diffusion model \citep{Ratcliff1978}, which has proven to provide
accurate descriptions of response times in many different settings.

One potential problem with the diffusion model is that it predicts a
continuous trade-off between speed and accuracy of responding, i.e.,
when participants are pressed to respond faster and faster, the
diffusion model predicts that this would lead to a gradual decrease in
accuracy.  The \code{speed} data set that we analyze below was
gathered to test this hypothesis versus the alternative hypothesis
stating that there is a sudden transition from slow and accurate
responding to fast responding at chance level.  At each trial of the
experiment, the participant is shown the current setting of the
relative reward for speed versus accuracy.  The bottom panel of
figure~ref{fig:speed} shows the values of this variable.  The main
question in this experiment was what would happen when this reward
variable would change from reward for accuracy only to reward for
speed only.  The \code{speed} data that we analyse here are from
participant A in Experiment 1 in \citet{Dutilh2009}, who provide a
complete description of the experiment and the relevant theoretical
background.

The central hypothesis about this data is whether indeed it is best 
described by two modes of responding rather than a single mode of 
responding with a continuous trade-off between speed and accuracy. 
The hallmark of having a discontinuity between slow versus speeded 
responding is that the switching between the two modes is assymetric
\citep[see e.g.][for a theoretical underpinning of this 
claim]{Maas1992}. The \code{fit} help page of \pkg{depmixS4} provides 
a number of examples in which the assymetricity of the switching 
process is tested; those examples are discussed at length in 
\citet{Visser2009b}.

@@

A dependent mixture model is defined by the number of states and the
initial state, state transition, and response distribution functions.
A dependent mixture model can be created with the
\code{depmix}-function as follows:
<<>>=
library(depmixS4)
data(speed)
set.seed(1)
mod <- depmix(response=rt~1, data=speed, nstates=2, trstart=runif(4))
@

The first line of code loads the \pkg{depmixS4} package and 
\code{data(speed)} loads the \code{speed} data set. The line
\code{set.seed(1)} is necessary to get starting values that will 
result in the right model, see more on starting values below.

The call to \code{depmix} specifies the model with a number of
arguments.  The \code{response} argument is used to specify the
response variable, possibly with covariates, in the familiar format
using formulae such as in \code{lm} or \code{glm} models.  The second
argument, \code{data=speed}, provides the \code{data.frame} in which
the variables from \code{response} are interpreted.  Third, the number
of states of the model is given by \code{nstates=2}.

\subsubsection{Starting values}

Note also that start values for the transition parameters are provided
in this call using the \code{trstart} argument.  At this time, the
package does not provide automatic starting values.  Starting values
for parameters can be provided using three arguments: \code{instart}
for the parameters relating to the prior probabilities,
\coce{trstart}, relating the transition models, and \code{respstart}
for the parameters of the response models.  Note that the starting
values for the initial and transition models as well as of the
multinomial logit response models are interpreted as {\em probabilities},
and internally converted to multinomial logit parameters.

\subsubsection{Fitting the model and printing results}

The \code{depmix} function returns an object of class \code{depmix}
which contains the model specification, and not a fitted model. Hence,
the model needs to be fitted by calling \code{fit}:

@@

The \code{summary} method of \code{fit}ted models provides the
parameter estimates, first for the prior probabilities model, second
for the transition models, and third for the response models.
<<>>=
summary(fm)
@

Since no further arguments were specified, the initial state, state
transition and response distributions were set to their defaults
(multinomial distributions for the first two, and a Gaussian
distribution for the response).

\subsection{Covariates on transition parameters\label{sec:trans}}

@@

\subsection{Adding covariates on the prior probabilities\label{sec:prior}}

To illustrate the use of covariates on the prior probabilities we have
included another data set with \pkg{depmixS4}.  The \code{balance}
data consists of 4 binary items (correct-incorrect) on a balance scale
task \citep{Siegler1981}.  The data form a subset of the data
published in \citet{Jansen2002}. Before specifying specifying a model 
for these data, we briefly describe them.

The balance scale taks is a famous task for testing cognitive
strategies developed by Piaget \citep{Siegler1981}.
Figure~\ref{fig:balance} provides an example of a balance scale item.
Participants' task is to say to which side the balance will tip when
released, or alternatively, whether it will stay in balance.  The item
shown in Figure~\ref{fig:balance} is a so-callled distance item: the
number of weights placed on each side is equal, and only the distance
of the weights to the fulcrum differs between each side.

\begin{figure}[htbp]
\begin{center}
	\includegraphic{graphs:baldist.pdf}
	\caption{Balance scale item; this is a so-called distance item.}
	\label{fig:balance}  
\end{center}
\end{figure}

Children in the lower grades of primary school are known to ignore the
distance dimension, and base their answer only on the number of
weights on each side.  Hence, they would typically provide the wrong
answer to these distance items.  Slightly older children do take
distance into account when responding to balance scale items, but they
only do so when the number of weights is equal on each side.  These
two strategies that children employ are known a Rule I and Rule II.
Other strategies can be teased apart by administering different items.
The \code{balance} data set that we analyse here consists of 4
distance items on a balance scale task administered to 779
participants ranging from 5 to 19 years of age.  The full set of items
consisted of 25 items; other items in the test are used to detect
other strategies that children (and young adults) employ in solving
balance scale items \citep[see][for details]{Jansen2002}.
